---
title: "CD34 vs H3K27M heatmaps"
author: "Giovani Genesi, MSc / Phillips Lab (UPenn)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
if(!require(ggplot2)){
install.packages("ggplot2")
}
if(!require(dplyr)){
install.packages("dplyr")
}

if (!require(profileplyr)) {
install.packages("profileplyr")
library(profileplyr)
}

if(!require(rtracklayer)){
BiocManager::install("rtracklayer")
}



# if(!require(ChIPseeker)){
# BiocManager::install("ChIPseeker")
# }

if(!require(tidytext)){
BiocManager::install("tidytext")
  library(tidytext)
}

# if(!require(TxDb.Hsapiens.UCSC.hg19.knownGene)){
# BiocManager::install("TxDb.Hsapiens.UCSC.hg19.knownGene")
# }

if (!require("RColorBrewer")) {
install.packages("RColorBrewer")
library(RColorBrewer)
}


knitr::opts_knit$set(root.dir = "C:/Users/thephillipslab/Documents/Projects/H3K27me3_CD34_cells/")
```

## Import signal quantification from deepTools as profileplyr object
```{r}
# EPZ_resistant_matrix <- import_deepToolsMat("./deepTools/EPZ6438_resistant_sites2.gz")
# EPZ_sensitive_matrix <- import_deepToolsMat("./deepTools/EPZ6438_sensitive_sites2.gz")

CD34vK27M <- import_deepToolsMat("./deepTools/EPZ6438_CD34_vs_H3K27M.gz")

```

The profileplyr object is a form of the RangedSummarizedExperiment class, which allows us to both store all of the relevant information that is imported from soGGi or deepTools, and have flexibility in how we manipulate this object in downstream analysis

```{r}
if(!require(SummarizedExperiment)){
BiocManager::install("SummarizedExperiment")
  library(SummarizedExperiment)
}

assays(EPZ_resistant_matrix)
assays(EPZ_sensitive_matrix) 

rowRanges(EPZ_sensitive_matrix)[1:3]
```

The information associated with each sample is stored as a Dataframe within with the parameters of the profileplyr object, and is accessed with sampleData().

```{r}
sampleData(EPZ_sensitive_matrix)[,1:6]
```

The ‘rowGroupsInUse’ element for the params slot indicates which column of the range metadata (mcols() or rowRanges()) that will be used for grouping in the final output if that object were used for visualization. For a profileplyr object created from a deepTools computeMatrix output, the inherited groups correspond to the BED files over which the signal was quantified, and these groups are contained in the ‘dpGroup’ column.

```{r}
params(EPZ_sensitive_matrix)$rowGroupsInUse

params(CD34vK27M)$rowGroupsInUse
```
```{r}
mcols(EPZ_resistant_matrix)$regions = factor("EPZ-resistant sites (CD34 cells)")
mcols(EPZ_sensitive_matrix)$regions = factor("EPZ-sensitive sites (CD34 cells)")
# params(EPZ_resistant_matrix)$rowGroupsInUse = mcols(EPZ_resistant_matrix)$regions
# params(EPZ_sensitive_matrix)$rowGroupsInUse = mcols(EPZ_sensitive_matrix)$regions

EPZ_sensitive_matrix <- groupBy(EPZ_sensitive_matrix, "regions")
params(EPZ_sensitive_matrix)$rowGroupsInUse

EPZ_resistant_matrix <- groupBy(EPZ_resistant_matrix, "regions")
```
```{r}
mcols(CD34vK27M)$Regions = NA
mcols(CD34vK27M)$Regions[mcols(CD34vK27M)$dpGroup == "H3K27me3_DMSO_subtract_EPZ6438.broadPeak"] = "EPZ-sensitive sites (CD34 cells)"
mcols(CD34vK27M)$Regions[mcols(CD34vK27M)$dpGroup == "H3K27me3_EPZ6438_SRR10957031.min500.gap2k.broad001.q001.mapQ30_peaks.broadPeak"] = "EPZ-resistant sites (CD34 cells)"

mcols(CD34vK27M)$Regions = factor(mcols(CD34vK27M)$Regions)

mcols(CD34vK27M)
```


The ‘mcolToOrderBy’ slot of the profileplyr object parameters indicates which column of the range metadata will be used for ordering the ranges as they are exported to either deepTools or to EnrichedHeatmap (with the generateEnrichedHeatmap() function within profileplyr). 

This can be set using the orderBy() function, which requires a profileplyr object and a character string matching a column name of the range metadata as arguments. If groupBy is never used on an object, the default will be to order by the mean signal of each range (within each group). 

In addition, until groupBy() has been used on a profileplyr object, the ‘mcolToOrderBy’ slot of the parameters will be NULL and will not be seen with params(proplyrObject).

```{r}
params(EPZ_sensitive_matrix)$mcolToOrderBy # defaults to NULL
```

To specify a column to reorder the ranges when they are exported:

```{r}
# EPZ_sensitive_matrix <- orderBy(EPZ_sensitive_matrix, "score")
# 
# params(EPZ_sensitive_matrix)$mcolToOrderBy # now will return "score"
# 
# EPZ_resistant_matrix <- orderBy(EPZ_resistant_matrix, "score")
```

## Subsetting the profileplyr object

The profileplyr object can be subset either by sample, or by rows and columns of the matrix for each sample. This is done using the ‘[ ]’ brackets, with the first position being assay matrix rows, the second position being assay matrix columns, and the third position being the entire matrix for each sample (in addition to the rest of the parameters and range data). 

```{r}
EPZ_sensitive_matrix[1:10,1:10] # returns the first 10 rows and first 10 columns of the matrix

# EPZ_sensitive_matrix[ , , 1:2] # returns the entire matrix for the first two samples
```

## Changing sample names

```{r}
# rownames(sampleData(EPZ_sensitive_matrix)) # returns sample names
# 
# rownames(sampleData(EPZ_sensitive_matrix)) <- "H3K27M cells" # reassigns samples' names
# 
# rownames(sampleData(EPZ_resistant_matrix)) <- "H3K27M cells"

rownames(sampleData(CD34vK27M)) <- "H3K27M cells"
```

# Add width of reagions 

```{r}
EPZ_sensitive_ranges = rowRanges(EPZ_sensitive_matrix) %>% ranges() %>% as.data.frame()

mcols(EPZ_sensitive_matrix)$widths = EPZ_sensitive_ranges$width
mcols(EPZ_sensitive_matrix)

EPZ_resistant_ranges = rowRanges(EPZ_resistant_matrix) %>% ranges() %>% as.data.frame()

mcols(EPZ_resistant_matrix)$widths = EPZ_resistant_ranges$width
mcols(EPZ_resistant_matrix)
```
# Export/Conversion of profileplyr object for heatmap visualization of ranges

To generate a heatmap within R directly from the profileplyr object, the generateEnrichedHeatmap() function should be used. This function takes a profileplyr object and produces a heatmap using the EnrichedHeatmap package. It allows for easy export from the profileplyr object and simple inclusion of the metadata as heatmap annotations. 

```{r}
# heatmap_EPZ_sensitive <- generateEnrichedHeatmap(EPZ_sensitive_matrix)
# heatmap_EPZ_resistant <- generateEnrichedHeatmap(EPZ_resistant_matrix)

heatmap_CD24 <- generateEnrichedHeatmap(CD34vK27M
                                        )


```

This function generates a multipanel heatmap with a variety of arguments that have been tailored to visualizing the profileplyr object. 

```{r}
class(heatmap_EPZ_resistant)
```

The convertToEnrichedHeatmapMat() function within profileplyr takes a profileplyr object as the only required input, and then converts the matrices contained within the assays slot to a list of ‘normalizedMatrix’ class objects that can be used in the EnrichedHeatmap() function. See the EnrichedHeatmap vignette for detailed examples for how heatmaps can be concatenated together to visualize all of the data in one figure.

```{r}
par(cex.lab=1.5, mfcol=c(2,2))
pdf("C:/Users/thephillipslab/Documents/Projects/H3K27me3_CD34_cells/images/CD34_vs_H3K27M.pdf",
    height = 6, width = 6)
heatmap_EPZ_sensitive 
heatmap_EPZ_resistant 
dev.off()
```